---
title: "Lung Cancer Prediction by Smoking and exposure to air pollution"
author: ""
format: html
execute: 
  echo: FALSE
  message: FALSE
---

*How does the risk of lung cancer vary across different levels of smoking and air pollution exposure?*

```{r}
#| label: setup
#| message: FALSE
library(tidyverse)
library(tidybayes)
library(brms)
library(ggthemes)
library(gtsummary)
```

```{r, message=FALSE, warning=FALSE, error=FALSE}
lung <- read_csv("data/lung-cancer.csv") |>
  rename(
    index = index,
    patient_id = `Patient Id`,
    age = Age,
    gender = Gender,
    air_pollution = `Air Pollution`,
    alcohol = `Alcohol use`,
    dust_allergy = `Dust Allergy`,
    occupational_hazards = `OccuPational Hazards`,
    genetic_risk = `Genetic Risk`,
    chronic_lung_disease = `chronic Lung Disease`,
    balanced_diet = `Balanced Diet`,
    obesity = Obesity,
    smoking = Smoking,
    passive_smoker = `Passive Smoker`,
    chest_pain = `Chest Pain`,
    coughing_blood = `Coughing of Blood`,
    fatigue = Fatigue,
    weight_loss = `Weight Loss`,
    shortness_of_breath = `Shortness of Breath`,
    wheezing = Wheezing,
    swallowing_difficulty = `Swallowing Difficulty`,
    clubbing_nails = `Clubbing of Finger Nails`,
    frequent_cold = `Frequent Cold`,
    dry_cough = `Dry Cough`,
    snoring = Snoring,
    level = Level
  ) |> mutate(gender = recode(gender, `1` = "Male", `2` = "Female"), level = recode(level, `Low` = 1, `Medium` = 2, `High` = 3))
```

```{r, message=FALSE, warning=FALSE, error=FALSE}
#| label: Gender Model
#| cache: TRUE

fit_sex <- brm(
  formula = level ~ gender,
  data = lung,
  family = categorical(),
  silent = 2,
  refresh = 0,
  seed = 9)
```

```{r, message=FALSE, warning=FALSE, error=FALSE}
#| label: Gender Model Plot 1

fit_sex |> add_epred_draws(newdata = tibble(gender = c("Female", "Male"))) |> mutate(.category = recode(.category, `1` = "Low", `2` = "Medium", `3` = "High")) |>
  select(gender, .category, .epred) |>
  filter(.category != "Medium") |>
  ggplot(aes(x = .epred, fill = gender)) +
    geom_histogram(bins = 100) +
    facet_grid(~ .category) +
    scale_x_continuous(
      breaks = c(0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1),
      labels = scales::percent_format()) +
  labs(title = "Male and Female Posterior for Risk of Lung Cancer", subtitle = "Men are twice as likely as women to get Lung Cancer", fill = "Gender", x = "Posterior Prediction for Lung Cancer") +
  theme(axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank())
```

## Project Summary

Lung cancer is the second most common cancer in the United States as of 2024. Drawing from a study conducted by Nature Medicine on over 462,000 individuals in China, who lived in areas with varying levels of air pollution, we aim to elucidate the relationship between lung cancer risk and air pollution. In our analysis, atmospheric conditions were treated as a categorical variable to assess their impact on lung cancer risk. Our findings indicate that individuals residing in more polluted environments are at a higher risk for lung cancer. Air pollution has a significantly greater effect on lung cancer risk compared to smoking, with an estimated difference in their impacts of 0.11 and a 95% credible interval ranging from 0.07 to 0.15.
