---
title: "Lung Cancer Prediction by Various Factors"
author: "Lowell Ethan Xavier"
format: html
execute: 
  echo: FALSE
  message: FALSE
---

*Modeling the relationships between lung cancer and other various factors such as air pollution, smoking, obesity, etc. to understand better treatment and prevention of it.*

```{r}
#| label: setup
#| message: FALSE
library(tidyverse)
library(tidybayes)
library(brms)
library(ggthemes)
library(gtsummary)
```

```{r, message=FALSE, warning=FALSE, error=FALSE}
lung <- read_csv("data/lung-cancer.csv") |>
  rename(
    index = index,
    patient_id = `Patient Id`,
    age = Age,
    gender = Gender,
    air_pollution = `Air Pollution`,
    alcohol = `Alcohol use`,
    dust_allergy = `Dust Allergy`,
    occupational_hazards = `OccuPational Hazards`,
    genetic_risk = `Genetic Risk`,
    chronic_lung_disease = `chronic Lung Disease`,
    balanced_diet = `Balanced Diet`,
    obesity = Obesity,
    smoking = Smoking,
    passive_smoker = `Passive Smoker`,
    chest_pain = `Chest Pain`,
    coughing_blood = `Coughing of Blood`,
    fatigue = Fatigue,
    weight_loss = `Weight Loss`,
    shortness_of_breath = `Shortness of Breath`,
    wheezing = Wheezing,
    swallowing_difficulty = `Swallowing Difficulty`,
    clubbing_nails = `Clubbing of Finger Nails`,
    frequent_cold = `Frequent Cold`,
    dry_cough = `Dry Cough`,
    snoring = Snoring,
    level = Level
  ) |> mutate(gender = recode(gender, `1` = "Male", `2` = "Female"), level = recode(level, `Low` = 1, `Medium` = 2, `High` = 3))
```

```{r, message=FALSE, warning=FALSE, error=FALSE}
#| label: Gender Model
#| cache: TRUE

fit_sex <- brm(
  formula = level ~ gender,
  data = lung,
  family = categorical(),
  silent = 2,
  refresh = 0,
  seed = 9)
```

```{r, message=FALSE, warning=FALSE, error=FALSE}
#| label: Gender Model Plot 1

fit_sex |> add_epred_draws(newdata = tibble(gender = c("Female", "Male"))) |> mutate(.category = recode(.category, `1` = "Low", `2` = "Medium", `3` = "High")) |>
  select(gender, .category, .epred) |>
  filter(.category != "Medium") |>
  ggplot(aes(x = .epred, fill = gender)) +
    geom_histogram(bins = 100) +
    facet_grid(~ .category) +
    scale_x_continuous(
      breaks = c(0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1),
      labels = scales::percent_format()) +
  labs(title = "Male and Female Posterior for Risk of Lung Cancer", subtitle = "Men are twice as likely as women to get Lung Cancer", fill = "Gender", x = "Posterior Prediction for Lung Cancer") +
  theme(axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank())
```

## Project Summary

Lung Cancer is the 2nd most common cancer affecting the United States in 2024. Using data from from a study done by the journal Nature Medicine on over 462,000 people in China who lived in areas with varying levels of air pollution, we seek to understand the relationship between risk of lung cancer and air pollution. We modeled people living in various atmospheric conditions, a categorical variable, as a categorical function of Lung Cancer Risk. Individuals with worse conditions were more likely to be prone to Lung Cancer. 
