---
title: "Lung Cancer Prediction by Various Factors"
author: "Lowell Ethan Xavier"
format: html
execute: 
    echo: FALSE
    message: FALSE
---

```{r}
#| label: setup
#| message: FALSE
library(tidyverse)
library(tidybayes)
library(brms)
library(ggthemes)
library(gtsummary)
```

```{r, message=FALSE, warning=FALSE, error=FALSE}
lung <- read_csv("data/lung-cancer.csv") |>
  rename(
    index = index,
    patient_id = `Patient Id`,
    age = Age,
    gender = Gender,
    air_pollution = `Air Pollution`,
    alcohol = `Alcohol use`,
    dust_allergy = `Dust Allergy`,
    occupational_hazards = `OccuPational Hazards`,
    genetic_risk = `Genetic Risk`,
    chronic_lung_disease = `chronic Lung Disease`,
    balanced_diet = `Balanced Diet`,
    obesity = Obesity,
    smoking = Smoking,
    passive_smoker = `Passive Smoker`,
    chest_pain = `Chest Pain`,
    coughing_blood = `Coughing of Blood`,
    fatigue = Fatigue,
    weight_loss = `Weight Loss`,
    shortness_of_breath = `Shortness of Breath`,
    wheezing = Wheezing,
    swallowing_difficulty = `Swallowing Difficulty`,
    clubbing_nails = `Clubbing of Finger Nails`,
    frequent_cold = `Frequent Cold`,
    dry_cough = `Dry Cough`,
    snoring = Snoring,
    level = Level
  ) |> mutate(gender = recode(gender, `1` = "Male", `2` = "Female"), level = recode(level, `Low` = 1, `Medium` = 2, `High` = 3))
```

Lung Cancer is the 2nd most common cancer affecting the United States in 2024, understanding the causes and predicting the probability of it can allow for the demystification and prevention of the disease. To do so, modelling the data is a great way to visualize the outcomes.

## Air Pollution + Smoking vs. Cancer Risk

```{r, message=FALSE, warning=FALSE, error=FALSE}
#| label: Air Pollution + Smoking Model
#| cache: TRUE
fit_pollute <- brm(
  formula = level ~ air_pollution,
  data = lung,
  family = categorical(), 
  refresh = 0, 
  silent = 2, 
  seed = 9)

fit_smoke <- brm(
  formula = level ~ smoking,
  data = lung,
  family = categorical(), 
  refresh = 0, 
  silent = 2, 
  seed = 9)
```

```{r, message=FALSE, warning=FALSE, error=FALSE}
#| label: New Data Creation
new_lung_air <- fit_pollute |> add_epred_draws(newdata = tibble(air_pollution = seq(1, 8, by = 1))) |> mutate(.category = recode(.category, `1` = "Low", `2` = "Medium", `3` = "High")) |> filter(!(.category %in% c("Medium")))
new_lung_smoke <- fit_smoke |> add_epred_draws(newdata = tibble(smoking = seq(1, 8, by = 1))) |> mutate(.category = recode(.category, `1` = "Low", `2` = "Medium", `3` = "High")) |> filter(!(.category %in% c("Medium")))
```

It is a common know fact that smoking has a direct correlation to lung cancer, and one of the greatest determinators of whether or not one can get lung cancer.

### Smoking

To find the relationship between lung cancer and smoking to see just by how much it increases risk of getting Lung Cancer, a categorical model can be made:

1. Formula
\begin{equation}
\text{level} = \alpha + \beta_1 (\text{smoking}) + \epsilon
\end{equation}

2. Table
```{r, message=FALSE, warning=FALSE, error=FALSE}
#| label: Smoking Regression

tbl_regression(fit_smoke)
```

3. Plot of Posterior Prediction (Air Pollution)
```{r, message=FALSE, warning=FALSE, error=FALSE}
#| label: Air Pollution + Smoking Model Plot 2

ggplot(new_lung_smoke, aes(x = as.factor(smoking), y = .epred, fill = .category)) +
  geom_dots(aes(color = .category), show.legend = c(color = FALSE)) +
  geom_line(aes(x = smoking, y = .epred, color = .category), show.legend = c(color = FALSE)) + 
  labs(x = "Smoking Level", y = "Predicted Cancer Risk", title = "Distribution of Predicted Cancer Risk by Smoking Level", fill = "Cancer Risk", subtitle = "The predicted cancer risk increases as the smoking level increases") +
  theme_clean() + 
  theme(
    legend.position = c(0.8, 0.5),
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 8),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major.x = element_line(color = "grey80", linetype = "dotted"),
    panel.grid.minor.x = element_line(color = "grey90", linetype = "dotted"))
```

As seen in the graph, the risk of lung cancer increases as the level of smoking increases, which is a well know fact, but something interesting happens when the prediction of lung cancer is seen with the level of air pollution.

### Air Pollution

To find the relationship between lung cancer and air pollution another categorical model can be made lke thus:

1. Formula
\begin{equation}
\text{level} = \alpha + \beta_1 (\text{air\_pollution}) + \epsilon
\end{equation}

2. Table
```{r, message=FALSE, warning=FALSE, error=FALSE}
#| label: Air Pollution Regression

tbl_regression(fit_pollute)
```

3. Plot of Posterior Prediction (Smoking)
```{r, message=FALSE, warning=FALSE, error=FALSE}
#| label: Air Pollution + Smoking Model Plot 1

ggplot(new_lung_air, aes(x = as.factor(air_pollution), y = .epred, fill = .category)) +
  geom_dots(aes(color = .category), show.legend = c(color = FALSE)) +
  geom_line(aes(x = air_pollution, y = .epred, color = .category), show.legend = c(color = FALSE)) + 
  labs(x = "Air Pollution Level", y = "Predicted Cancer Risk", title = "Distribution of Predicted Cancer Risk by Air Pollution Level", fill = "Cancer Risk", subtitle = "The predicted cancer risk increases as the air pollution level increases") +
  theme_clean() + 
  theme(
    legend.position = c(0.8, 0.5),
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 8),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major.x = element_line(color = "grey80", linetype = "dotted"),
    panel.grid.minor.x = element_line(color = "grey90", linetype = "dotted"))
```

Comparing both posterior prediction graphs we can see that they are similiar in terms of trend and probability of risk of lung cancer, leading to an interesting revelation that...

> Air pollution level is as bad as smoking for probability of getting lung cancer

